Normalization Circularity in Population-Based FA Analysis

### 2011-06-06 ANTsTBSS/paper.tex
ENH: compile script + revised abstract

\begin{abstract}

%% Text of abstract
Critical to many studies investigating white matter differences across populations 
is normalization of fractional anisotropy (FA) images to a common coordinate system
for statistical analysis.  Although variations exist, common to several current 
approaches is direct FA-to-FA registration whether it be to a chosen individual or
a generated template.  This type of circular analysis 
is susceptible to the general confounding 
effect of minimizing data differences with nonrigid registration which are potentially 
the precise differences to be statistically quantified.  We demonstrate that
population normalization to a template explicitly optimizes certain voxelwise statistical 
testing measures as opposed to maximizing anatomical alignment.  
As a corrective, we advocate the use of the Symmetric Group Normalization strategy for 
unbiased template construction from anatomical (e.g. T1-weighted) images which provides 
the coordinate system for statistical inference of FA differences.
Following template construction, individual FA images are warped to the template via 
the corresponding anatomical image.  Performance is evaluated using open data from the International 
Neuroimaging Data-sharing Initiative and a cohort of 16 traumatic brain injury survivors compared with 
17 control subjects.  Alignment comparisons with the normalization component of the 
popular tract-based spatial statistics (TBSS)
show that our anatomically-based template approach produces good alignments despite
the absence of any direct FA-to-FA registration.  Of practical consideration we note that the 
software tools discussed in this work are provided to the public as part of the open source 
Advanced Normalization Tools (ANTs) repository.

@@

>>>>

\begin{abstract}
Template-based normalization plays a critical role in many population
studies based on diffusion tensor imaging (DTI).  Several current
approaches compute the fractional anisotropy (FA) from the DTI and then
directly register the FA image to a FA template before performing
statistical comparison between different groups within a population.  In
this paper, we demonstrate that this analysis strategy induces selection
bias that varies in strength with the similarity metric used in
registration.  Specifically, we show that the popular sum of squared
difference intensity metric maximizes effect size, rather than
anatomical alignment, and consistently overestimates statistical
significance.  Consequently, we advocate a more traditional approach
that normalizes a population of FA images through the associated T1
anatomical images.  We show that the Symmetric Group Normalization
strategy for unbiased template construction from anatomical (e.g.
T1-weighted) images provides a valid and unbiased coordinate system for
statistical inference of FA differences and yet retains detection power
in comparison to FA-based normalization methods.  We establish this
outcome by using open data in normal controls from the International
Neuroimaging Data-sharing Initiative as well as a cohort of 16 traumatic
brain injury survivors compared with 17 control subjects.  Alignment
comparisons with the normalization component of the popular tract-based
spatial statistics (TBSS) show that our anatomically-based template
approach produces good alignments despite the absence of any direct
FA-to-FA registration.  Of practical consideration we note that the
software tools discussed in this work are provided to the public as part
of the open source Advanced Normalization Tools (ANTs) repository.

@@

Normalization (also known as image registration) maps each image in the population to a common coordinate system thus enabling statistical comparison across patient groups or association with other variables such as genetics, cognition or age.
